Bifurcation or state tipping: assessing transition type in a model trophic cascade

### 2018-07-14 paper/paper.Rmd
add 4 figures

![S-tipping vs B-tipping.  A critical transition in a system with alternative stable states can be driven either by a perturbation to the state, driving the system across the separatrix (top panel), or by a bifurcation, in which a stable point is annihilated (bottom panel)](figures/scheffer-fig.png)

@@

Central to resolving this question is untangling a second issue that frequently arises in the discussion of alternative stable state dynamics in ecological systems: the issue of dimension.  The S-tipping and B-tipping transitions illustrated in Figure 1 are fundamentally rooted in one-dimensional analysis: the state consists of a single continuously valued dynamical variable (i.e. population density of adult bass).  Ecosystems are notoriously more complex, and this lake ecosystem is no exception. Here, we use analyze previously published models  of trophic cascades to determine whether manipulating the top predator constitutes B-tipping or S-tipping. Common versions include 5 state variables, ranging from top predators to primary producers. We accomplish this by performing dimension reduction on the full 5D model, and surmising whether the reduced model can be expressed in a way such that manipulating predator abundance is tantamount to manipulating a parameter, or if it necessarily constitutes state manipulation. This exercise facilitates the interpretation of the lake experiment in the context of predicting regime shifts by using a 1D model to understand a high dimensional system.

@@

These same dynamics can be seen in the direct numerical simulations of the full 5-dimensional model.  At a q of 0, the fish in the simulation underwent a critical transition (Figure 2). This transition occurred during time step 510, although an abrupt change in state was not observed until step 850. The changes in A and J were similar, but were inversely related to those in F. The predation effect of A on J was small compared to the role of A's fecundity in producing J and A's role as a predator of F, which negatively impacts J as well. Overall, both bass time series decreased with increasing q (abruptly when stocking converted to harvest at q=0), whereas the abundance of planktivorous fish was positively correlated with $q$.

@@

These results demonstrate that the system considered can indeed be driven through fold bifurcations as the parameter $q$ is slowly tuned up or down, forcing the creation or annihilation of an alternative stable state -- confirming the presence of B-tipping in this system. Simulations of the same system using previous estimates for ecological parameters confirm that the predictions of the 1-dimensional approximation (eq 9) correspond with the dynamics observed in the full model.

>>>>

![Figure 1: S-tipping vs B-tipping.  A critical transition in a system with alternative stable states can be driven either by a perturbation to the state, driving the system across the separatrix (top panel), or by a bifurcation, in which a stable point is annihilated (bottom panel)](figures/scheffer-fig.png)

@@

Central to resolving this question is untangling a second issue that frequently arises in the discussion of alternative stable state dynamics in ecological systems: the issue of dimension.  The S-tipping and B-tipping transitions illustrated in Figure 1 are fundamentally rooted in one-dimensional analysis: the state consists of a single continuously valued dynamical variable (i.e. population density of adult bass).  Ecosystems are notoriously more complex, and this lake ecosystem is no exception. Here, we use analyze previously published models  of trophic cascades to determine whether manipulating the top predator constitutes B-tipping or S-tipping. Common versions include 5 state variables, ranging from top predators to primary producers (Figure 2). We accomplish this by performing dimension reduction on the full 5D model, and surmising whether the reduced model can be expressed in a way such that manipulating predator abundance is tantamount to manipulating a parameter, or if it necessarily constitutes state manipulation. This exercise facilitates the interpretation of the lake experiment in the context of predicting regime shifts by using a 1D model to understand a high dimensional system.

![Figure 2: Food Web Heuristic.  Adult bass give rise to juveniles, while eating their own young and pumpkinseed sunfish. Sunfish compete with and eat juvenile bass. Sunfish eat herbivorous zooplankton like *Daphnia*, which in turn eat phytoplankton. ](figures/Slide1.png)

@@

![Figure 3. Time series of the 5 state variables, with q (harvest rate of adult bass) being increased linearly over time. ](../pkgBuild/manuscript_report/figure-5dsimulation-1.png)

These same dynamics can be seen in the direct numerical simulations of the full 5-dimensional model.  At a q of 0, the fish in the simulation underwent a critical transition (Figure 4). This transition occurred during time step 510, although an abrupt change in state was not observed until step 850. The changes in A and J were similar, but were inversely related to those in F. The predation effect of A on J was small compared to the role of A's fecundity in producing J and A's role as a predator of F, which negatively impacts J as well. Overall, both bass time series decreased with increasing q (abruptly when stocking converted to harvest at q=0), whereas the abundance of planktivorous fish was positively correlated with $q$.

![Figure 4. Top row: motion of A for different values of q. Bottom row: the negative integral of the top row, depecting in the so-called ball-in-cup diagrams.](../pkgBuild/manuscript_report/figure-Amotion-ballincup-1.png)

@@

These results demonstrate that the system considered can indeed be driven through fold bifurcations as the parameter $q$ is slowly tuned up or down (Figure 5), forcing the creation or annihilation of an alternative stable state -- confirming the presence of B-tipping in this system. Simulations of the same system using previous estimates for ecological parameters confirm that the predictions of the 1-dimensional approximation (eq 9) correspond with the dynamics observed in the full model.

![Figure 5. Bifurcation diagram derived from eq 9. Solid lines are stable equilibria, the dashed line is the unstable equilibrium or separatrix.](../pkgBuild/manuscript_report/figure-bifurcationDiagram-1.png)
